---
title: "Workshop 1: QCBS R Workshop Series"
author: Québec Centre for Biodiversity Science
output:
  revealjs::revealjs_presentation:
    css: custom-qcbs.css
    fig_caption: yes
    theme: white
    highlight: haddock
    center: true
    transition: fade
    background_transition: fade
    self_contained: false
    pandoc_args: [ "--slide-level", "2" ]
    reveal_plugins: ["notes", "search", "zoom"]
---


# Using R as a calculator

## Addition, substraction, etc.

- Additions and Substractions

```{r}
	1+1
  10-1
```

- Multiplications and Divisions

```{r]
	2*2
  8/2
```

- Exponents

```{r}
	2^3
```

## Challenge 2

Use R to calculate the following skill testing question:

$$
2+16*24-56
$$

<small>Hints: The `*` symbol is used to multiply</small>

## Challenge 2: Solution

## Challenge 2: Solution

- Question

$$
2+16*24-56
$$

- Solution

```{r}
	2+16*24-56
```

## Challenge 3

Use R to calculate the following skill testing question:

$$
2+16*24-56 / (2+1)-457
$$

<small>Hints: Think about the order of the operation.</small>

## Challenge 3: Solution

- Question

$$
2+16*24-56 / (2+1)-457
$$

- Solution

```{r}
	2+16*24-56/(2+1)-457
```

Note that R follows the order of the operations

## R command line tip

Use the "Up" and "Down" arrow keys to reproduce previous commands.

Give it a try!

## Challenge 4

<div id="right">


```{r echo = F, fig.width=5, fig.height=5}
	plot(c(-1, 1), c(-1, 1), type = "n", ann = F, axes = F, asp = 1)
	lines(x = radius * cos(theta), y = radius * sin(theta))
	arrows(0, 0, x1 = 1, length = 0)
	text(.5, .1, "radius = 5 cm", cex = .8)
```

</div>

<div id="left">

What is the area of this circle, with a radius of $5\ cm$?

</div>

## Challenge 4: Solution

- Question

What is the area of a circle, with a radius of $5\ cm$?

- Solution

```{r}
  3.1416*5^2
```

Tip: Note that `R` has some built-in constant such as $\pi$, so you can write:

```{r}
  pi*5^2
```

# Objects and indexing

## Objects

- One of the most useful concept in `R`!
- You can store values as named objects using the assignement operator `<-`

```{r, eval = F}
  object name <- assigned value
```

The value on the **right** is assigned to the name on the **left** with the assignement operator `<-`

 <small>It is also possible to use the `=` sign, but it is better to **avoid it** as it is also used for other purposes.</small>
 
## Objects names

test 2
